Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2017-03-21 paper/paper-v3.tex
revision new sim setup

Communicative efficiency is sensitive to $\lambda$ as it influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent. As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states less equivocally fare better. This difference increases with $\lambda$ as it promotes more rational linguistic behavior, widening the gap in expressivity between some types. As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics under any $\lambda$-value. For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together. %Across types and populations incumbents make up $.50$ of the mean outcome.

@@

The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others such as $\Lbound$ in virtue of the relatively simple semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different observable behavior because its pragmatic use approximates a one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$, learnability does not succeed in selecting for a single type of a kind. For instance, for only one of the six pragmatic $\Llack$ types. Instead, it leads to their coexistence. Each is passed on to next generations with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other.

As implicitly suggested by Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality in linguistic behavior, with each reaching approximately the same proportion of users in the population. Low values of $\lambda$ level the differences in observable behavior across types and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as literal $\Lall$. A pressure for learnability alone may consequently lead to the selection of communicatively suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.

\subsubsection{Expressivity and learnability}
The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for a division of labor between semantics and pragmatics. Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers. When pressured toward expressivity, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection. When pressured toward learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more complex alternatives such as $\Lbound$. However, on its own learnability does not foment the propagation of these types across the population. What is more, it leads to heterogeneous populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$ types.

@@

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values. These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics paired with pragmatic language use. The proportion of these types increases with $\lambda$ and $l$. Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen meaning more likely to be adopted over alternatives that do not. In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.

@@

To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors. First, the degree to which linguistic behavior is rational in that agents try to maximize communicative success. This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others. Second, the inductive bias, which leads learners to prefer simpler over more complex semantic representations in acquisition. Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.

@@

The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning. The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model expressivity as exerting its force only in the production of learning data. The degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is absent. That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence. Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased with $\Lall$. This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}). In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to communicatively distinguish meanings more accurately  (e.g., \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}). It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants. However, on a general level they may suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer. That is, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.

\section{Conclusion}
The cultural evolution of meaning is influenced by intertwined pressures. We set out to investigate this process by putting forward a model that combines a pressure toward successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition. Its objects of selection and replication are pairs of lexical meanings and patterns of language use. This allows the model to trace the interaction between conventional meaning and pragmatic use. Additionally, it takes the challenge serious of neither semantics nor pragmatics being directly observable. Instead, learners need to infer these unobservables from overt data that results from their combination.  These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use. That is, the relative learning advantage of simpler semantics in combination with functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.

>>>>

Expressivity is sensitive to $\lambda$ as it influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent. Low $\lambda$ levels communicative differences between types, as even those that can differentiate between all states semantically or pragmatically do not always maximize expected communicative success. The outcome is that of a mixed population that allows for the survival of competing types. Conversely, higher $\lambda$ promotes more rational linguistic behavior, widening the gap in expressivity between types and promoting more homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a lack of upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together. %Across types and populations incumbents make up $.50$ of the mean outcome.

@@

The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different observable behavior because its pragmatic use approximates a one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$, learnability alone does not succeed in selecting for a single type of a kind. For instance, for only one of the six pragmatic $\Llack$ types. Instead, it leads to their coexistence. Each is passed on to next generation with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$, with each reaching approximately the same proportion of users in the population. As with expressivity only, low values of $\lambda$ make the differences in observable behavior across types less pronounced and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.

\subsubsection{Expressivity and learnability}
The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for the emergence of a division of labor between semantics and pragmatics. Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers. When pressured toward expressivity, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection. When pressured toward learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more complex alternatives such as $\Lbound$. However, on its own learnability does not foment the propagation of these types across the population. What is more, it leads to heterogeneous populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$ types.

@@

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values. These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics paired with pragmatic language use. The proportion of these types increases with $\lambda$ and $l$. Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen lexical meaning more likely to be adopted over alternatives that do not. %In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.

@@

To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors. First, the degree to which agents try to maximize communicative success. This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others. Second, the inductive bias, which leads learners to prefer simpler over more complex semantic representations in acquisition. Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.

@@

The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, learning involving a joint inference over types of pragmatic behavior and lexical meaning, and, accordingly, the possibility to trace the co-evolution of conventional semantics and pragmatic use. The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model expressivity as exerting its force only in the production of learning data. The degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is absent. That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al. only consider the latter's influence. Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by $\Lall$. This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}). In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to communicatively distinguish meanings more accurately  (e.g., \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}). It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants. However, on a general level they may suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer. That is, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.

\section{Conclusion}
The cultural evolution of meaning is influenced by intertwined pressures. We set out to investigate this process by putting forward a model that combines a pressure toward successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition. Its objects of selection and replication are pairs of lexical meanings and patterns of language use. This allows the model to trace the interaction between conventional meaning and pragmatic use. Additionally, it takes the challenge serious of neither semantics nor pragmatics being directly observable. Instead, learners need to infer these unobservables from overt data that results from their combination.  These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use. That is, the relative learning advantage of simpler semantics in combination with functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, to a division of labors between semantics and pragmatics.
